---
title: "p8105_hw2_khs2318"
author: "Kayla Schiffer-Kane"
date: "2023-10-04"
output: 
  github_document:
    toc: true
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(message = FALSE)

knitr::opts_chunk$set(message = FALSE,
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)

library(p8105.datasets)
library(tidyverse)
library(ggridges)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```



# Problem 1

## Load dataset
```{r load_instacart}

data("instacart")
insta_df = instacart 
#skimr::skim(insta_df)
#length(unique(insta_df$aisle_id)) # 134 
```

## Describe dataset 

**Write a short description of the dataset, noting the size and structure of the data, describing some key variables, and giving illustrative examples of observations.**

The dataset has `r nrow(insta_df)` rows and `r ncol(insta_df)` columns. Each row represents a product with a unique ID and details of its order (day and time), as well as details of the product. `add_to_cart_order` is a numerical variable that indicates the order in which items were added to the cart. `aisle_id` and `department_id` are similar numeric variables that indicate which aisle and department the product is found in. There are also product names (e.g., Bulgarian Yogurt, Organic 4% Milk Fat Whole Milk Cottage Cheese), aisle names (e.g., yogurt, other creams cheeses), and department names (e.g., dairy eggs, produce). 

**How many aisles are there, and which aisles are the most items ordered from?** There are `r length(unique(insta_df$aisle_id))` aisles, and the most items are ordered from fresh vegetables and fresh fruits

```{r aisles_count}
ailse_ct = 
  insta_df |>
  group_by(aisle) |>
  summarize(n_obs = n()) |>
  arrange(desc(n_obs))
```


**Make a plot that shows the number of items ordered in each aisle, limiting this to aisles with more than 10000 items ordered. Arrange aisles sensibly, and organize your plot so others can read it.** 

```{r aisles_plot}
ailse_ct |>
  filter(n_obs > 10000) |> 
  mutate(aisle = fct_reorder(aisle, n_obs)) |> 
  ggplot(aes(x = aisle, y = n_obs)) + 
  geom_point() + 
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```


**Make a table showing the three most popular items in each of the aisles “baking ingredients”, “dog food care”, and “packaged vegetables fruits”. Include the number of times each item is ordered in your table.**
```{r aisle_popular}
insta_df |>
  filter(aisle %in% c("baking ingredients","dog food care","packaged vegetables fruits")) |>
  group_by(aisle, product_name) |>
  summarize(n_obs = n()) |>
  filter(min_rank(desc(n_obs)) < 4) |>
  arrange(desc(n_obs)) |>
  knitr::kable()
```

**Make a table showing the mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are ordered on each day of the week; format this table for human readers (i.e. produce a 2 x 7 table).**
```{r mean_hour}
insta_df |>
  filter(product_name %in% c("Pink Lady Apples","Coffee Ice Cream")) |>
  group_by(product_name, order_dow) |>
  summarize(mean_hour = mean(order_hour_of_day)) |> 
  pivot_wider(names_from = order_dow,
              values_from = mean_hour) |>
  knitr::kable()

```


# Problem 2

First, do some data cleaning:

format the data to use appropriate variable names;
focus on the “Overall Health” topic
include only responses from “Excellent” to “Poor”
organize responses as a factor taking levels ordered from “Poor” to “Excellent”

```{r load_brfss}
library(p8105.datasets)
data("brfss_smart2010")
brfss_data = 
  brfss_smart2010 |>
  janitor::clean_names() |>
  rename(location_abbr = locationabbr,  location_desc = locationdesc, 
         resp_id = respid) |>
  filter(topic == "Overall Health" & response %in% c("Excellent", "Poor")) |>
  mutate(response = factor(response, levels = c("Poor", "Excellent"), ordered = TRUE))
skimr::skim(brfss_data)


```




