---
title: "p8105_hw2_khs2318"
author: "Kayla Schiffer-Kane"
date: "2023-10-04"
output: 
  github_document:
    toc: true
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(message = FALSE)

knitr::opts_chunk$set(message = FALSE,
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)

library(p8105.datasets)
library(tidyverse)
library(ggridges)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```



# Problem 1

## Load dataset
```{r load_instacart}

data("instacart")
insta_df = instacart 
#skimr::skim(insta_df)
#length(unique(insta_df$aisle_id)) # 134 
```

## Describe dataset 

**Write a short description of the dataset, noting the size and structure of the data, describing some key variables, and giving illustrative examples of observations.**

The dataset has `r nrow(insta_df)` rows and `r ncol(insta_df)` columns. Each row represents a product with a unique ID and details of its order (day and time), as well as details of the product. `add_to_cart_order` is a numerical variable that indicates the order in which items were added to the cart. `aisle_id` and `department_id` are similar numeric variables that indicate which aisle and department the product is found in. There are also product names (e.g., Bulgarian Yogurt, Organic 4% Milk Fat Whole Milk Cottage Cheese), aisle names (e.g., yogurt, other creams cheeses), and department names (e.g., dairy eggs, produce). 

**How many aisles are there, and which aisles are the most items ordered from?** There are `r length(unique(insta_df$aisle_id))` aisles, and the most items are ordered from fresh vegetables and fresh fruits

```{r aisles_count}
ailse_ct = 
  insta_df |>
  group_by(aisle) |>
  summarize(n_obs = n()) |>
  arrange(desc(n_obs))
```


**Make a plot that shows the number of items ordered in each aisle, limiting this to aisles with more than 10000 items ordered. Arrange aisles sensibly, and organize your plot so others can read it.** 

```{r aisles_plot}
ailse_ct |>
  filter(n_obs > 10000) |> 
  mutate(aisle = fct_reorder(aisle, n_obs)) |> 
  ggplot(aes(x = aisle, y = n_obs)) + 
  geom_point() + 
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```


**Make a table showing the three most popular items in each of the aisles “baking ingredients”, “dog food care”, and “packaged vegetables fruits”. Include the number of times each item is ordered in your table.**
```{r aisle_popular}
insta_df |>
  filter(aisle %in% c("baking ingredients","dog food care","packaged vegetables fruits")) |>
  group_by(aisle, product_name) |>
  summarize(n_obs = n()) |>
  filter(min_rank(desc(n_obs)) < 4) |>
  arrange(desc(n_obs)) |>
  knitr::kable()
```

**Make a table showing the mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are ordered on each day of the week; format this table for human readers (i.e. produce a 2 x 7 table).**
```{r mean_hour}
insta_df |>
  filter(product_name %in% c("Pink Lady Apples","Coffee Ice Cream")) |>
  group_by(product_name, order_dow) |>
  summarize(mean_hour = mean(order_hour_of_day)) |> 
  pivot_wider(names_from = order_dow,
              values_from = mean_hour) |>
  knitr::kable()
```


# Problem 2

## Load and Clean Dataset

* Format the data to use appropriate variable names;
* Focus on the “Overall Health” topic
* Include only responses from “Excellent” to “Poor”
* Organize responses as a factor taking levels ordered from “Poor” to “Excellent”

```{r load_brfss}
library(p8105.datasets)
data("brfss_smart2010")
brfss_data = 
  brfss_smart2010 |>
  janitor::clean_names() |>
  rename(state = locationabbr,  county = locationdesc, 
         resp_id = respid) |>
  filter(topic == "Overall Health" & response %in% c("Excellent", "Poor")) |>
  mutate(response = factor(response, levels = c("Poor", "Excellent"), ordered = TRUE))
```

## Data Analysis 

### Locations by State 

**In 2002, which states were observed at 7 or more locations? What about in 2010?**

* Because `location_id` are all NA, using `geo_locaiton` as the best measure of different locations. County could also be used, but some locations may be within a single county. 

```{r states_exp}
locations_2002 = 
  brfss_data |>
  filter(year == 2002) |>
  group_by(state) |>
  summarize(
    n_obs = n(),
    n_location = n_distinct(geo_location)) |> 
  filter(n_location >= 7)
head(locations_2002, n=5)

locations_2010 = 
  brfss_data |>
  filter(year == 2010) |>
  group_by(state) |>
  summarize(
    n_obs = n(),
    n_location = n_distinct(geo_location)) |> 
  filter(n_location >= 7)
head(locations_2010, n=5)

```

There are `r nrow(locations_2002)` states observed at 7 or more locations in 2002 (including Connecticut, Florida, and Massachusetts), and `r nrow(locations_2010)` in 2010 (including California, Colorado, and Florida.

### "Excellent" Responses

**Construct a dataset that is limited to Excellent responses, and contains, year, state, and a variable that averages the data_value across locations within a state.**

```{r exc_dataset}
exc_dataset = 
  brfss_data |>
  filter(response == "Excellent") |>
  group_by(year, state) |>
  summarize(n_obs = n(), 
            data_average = mean(data_value))
head(exc_dataset)

```

**Make a “spaghetti” plot of this average value over time within a state (that is, make a plot showing a line for each state across years – the geom_line geometry and group aesthetic will help).**
```{r spaghetti_plot}
exc_dataset |>
  ggplot(aes(x = year, y = data_average)) + 
  geom_line(aes(group = state, colour = state)) 

```


### NY state: 2006, 2010
**Make a two-panel plot showing, for the years 2006, and 2010, distribution of data_value for responses (“Poor” to “Excellent”) among locations in NY State.**
```{r two_panel_NY}
ny_dataset = 
  brfss_data |>
  filter(year %in% c(2006, 2010) & state == 'NY') |>
  mutate(year = as.factor(year))

ny_dataset |>
  ggplot(aes(x = data_value, fill = year)) + 
  geom_histogram(position = "dodge", binwidth = 0.5) + 
  facet_grid(.~year)

```

# Problem 3

## Load and Tidy NHANES Datasets

**Load, tidy, merge, and otherwise organize the data sets. Your final dataset should include all originally observed variables; exclude participants less than 21 years of age, and those with missing demographic data; and encode data with reasonable variable classes (i.e. not numeric, and using factors with the ordering of tables and plots in mind).**

```{r load_nhanes_data}
demo_df = 
  read_csv('./data/nhanes_covar.csv', skip = 4) |>
  janitor::clean_names() |>
  mutate(sex = recode(sex, "2" = "Female", "1" = "Male")) |>
  mutate(sex = factor(sex, levels = c("Male", "Female"), ordered = TRUE)) |>
  mutate(education = recode(education, "1" = "hs_less", "2" = "hs_equiv", "3" = "hs_more")) |>
  mutate(education = factor(education, levels = c("hs_less", "hs_equiv","hs_more"), ordered = TRUE)) |>
  filter(age >= 21) |> 
  filter(!is.na(sex) & !is.na(age) & !is.na(education) & !is.na(bmi)) 

accel_df =
  read_csv('./data/nhanes_accel.csv') |>
  janitor::clean_names()
```

### Merge NHANES 
```{r merge_nhanes}
nhanes_df = 
  left_join(demo_df, accel_df, by = "seqn")
head(nhanes_df)
```


## Men and Women, Education and Age
**Produce a reader-friendly table for the number of men and women in each education category, and create a visualization of the age distributions for men and women in each education category. Comment on these items.**

```{r education_table}
nhanes_df |>
  group_by(education, sex) |>
  summarize(n_indiv = n()) |>
  knitr::kable()

nhanes_df |>
  ggplot(aes(x = age, fill = education)) +
  geom_density(alpha = 0.4, adjust = 0.5) +
  facet_grid(.~ sex)

```


There is a roughly even split of Male versus Female for education levels of less than or more than high school. For high school equivalent, there are more Males (`r (35/(35 + 23)*100) |> round(2)`%) than Females (`r (23/(35 + 23)*100) |> round(2)`%). For both Males and Females, but for Females in particular, participants with more than high school education are moslty young (< 40). For Females in particular, the older participants (>60) seem to have more concentration of less than or equivalent to high school education. The distribution of Males with less than high swchool education seem to be largely beteen ages 40-50 or greater than 60. 

## Accelemator Analyses

**Traditional analyses of accelerometer data focus on the total activity over the day. Using your tidied dataset, aggregate across minutes to create a total activity variable for each participant. Plot these total activities (y-axis) against age (x-axis); your plot should compare men to women and have separate panels for each education level. Include a trend line or a smooth to illustrate differences. Comment on your plot.**

```{r}
total_activity = 
  nhanes_df |>
  pivot_longer(min1:min1440, 
               names_to = "min",
               values_to = "min_value") |>
  group_by(seqn, age, sex, education) |>
  summarize(n_obs = n(),
            total_activity = sum(min_value))

total_activity |>
  ggplot(aes(x = age, y = total_activity, color = sex)) + 
  geom_point() + 
  geom_smooth(se = FALSE) + 
  facet_grid(. ~ education)
```

For less than high school level education, total acitvity on average is higher for Females below the age of 40, and higher for Males above the age of 40. For high school equivalent or more education level, total activity is on average higher for women than men for all ages (with the exception of a similar trend in activity levels for Males and Females around the age of 20 for high school equivalent). For all education levels, activity level decreases with age, but the decrease with age is less severe with increased education level (that is, for less than high school education, the dip in activity with increased age is quite drastic, and less so for high school or more).


## Daily Accelerometer
Accelerometer data allows the inspection activity over the course of the day. Make a three-panel plot that shows the 24-hour activity time courses for each education level and use color to indicate sex. Describe in words any patterns or conclusions you can make based on this graph; including smooth trends may help identify differences.



